---
title: GLS, MMRM, and Power
date: "2014-04-15"
categories: [mixed-effects, MMRM, gls, lme, nlme, R, power, simulation]
summary: Analytic and simulated power for MMRM with missing data.
---
# Power for MMRM

We use the power calculation formula of [Lu, Luo and Chen (2008)](http://dx.doi.org/10.2202/1557-4679.1098) for Mixed Models of Repeated Measures, which accomodates general missing data patterns. This formula is implemented in the [longpower](http://cran.r-project.org/web/packages/longpower/index.html) package. We also verify the formula with a simple simulation.

## Load packages

```{r library, results='hide', message=FALSE}
library(longpower)
library(nlme)
library(mvtnorm)
library(parallel)
library(ggplot2)
library(contrast)
```

```{r, results='hide', echo=FALSE}
knitr::opts_chunk$set(fig.width=1.7*7, fig.height=7, crop=TRUE)
theme_set(theme_bw(base_size=12))
# The palette with black:
cbbPalette <- c("#0072B2", "#a21933", "#CC79A7", "#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#D55E00")
dxpal <- c("#2b8cbe", "#fc8d59","#e34a33", "#fdcc8a", "#74a9cf")
Palette0 <- Palette <- dxpal
scale_colour_discrete <- function(...) scale_colour_manual(..., values=cbbPalette)
scale_fill_discrete <- function(...) scale_fill_manual(..., values=cbbPalette)
```

## Simulation settings
```{r sim_settings}
gls_sigma <- 2
gls_weights <- c(1, 1.5, 1.75, 2.0)
placebo_mean <- c(0, 0, 0, 0)
active_mean <- c(0, 1, 1.5, 1.8)
retention <- c(1, 0.9, 0.8, 0.7)
# compound symmetric R
R <- matrix(0.5, nrow = 4, ncol = 4)
diag(R) <- 1
N <- 200
```

## Analytic calculations
```{r calcs}
power.mmrm(power=0.80, ra = retention, Ra=R, N=N, sigmaa = gls_sigma*gls_weights[4])
power.t.test(power=0.80, n = N/2*retention[4], sd = gls_sigma*gls_weights[4])
```

The t-test delta is larger, which is expect given the conservative handling of the assumed retention rate. The MMRM makes use of partial informatin from dropouts.

## Set up data for simulation

```{r, sim_data_setup}
dd0 <- expand.grid(time = as.factor(1:4), id = 1:N)
dd0$Time <- as.numeric(dd0$time)
dd0$trt <- ifelse(dd0$id <= N/2, 0, 1)
dd0 <- as.data.frame(model.matrix(~ -1+id+trt*time+Time, data = dd0))
dd0[,'trt:time1'] <- with(dd0, trt*time1)
dd0$Trt <- as.factor(dd0$trt)
dd0$time <- as.factor(dd0$Time)
dd0$Y0[dd0$trt==1] <- as.matrix(subset(dd0, trt==1, c('trt:time1', 'trt:time2', 
  'trt:time3', 'trt:time4'))) %*% active_mean
dd0$Y0[dd0$trt==0] <- as.matrix(subset(dd0, trt==0, c('time1', 'time2', 
  'time3', 'time4'))) %*% placebo_mean
ggplot(dd0, aes(x=Time, y=Y0, color=Trt)) + geom_line()

set.seed(20130318)
dd1 <- merge(dd0, data.frame(id = 1:N, drop = sample(1:5, N, replace=TRUE, p=c(0,0.1,0.1,0.1,0.7))))
dd1$e <- as.numeric(t(rmvnorm(n=N, mean=rep(0,4), sigma=gls_sigma^2*diag(gls_weights)%*%R%*%diag(gls_weights))))
dd1$Y <- with(dd1, Y0 + e)
dd1 <- subset(dd1, Time < drop)
ggplot(dd1, aes(x=Time, y=Y, color=Trt, group = id)) + geom_line()

fit0 <- gls(Y~-1+(time1+time2+time3+time4)+(time1+time2+time3+time4):trt, 
    correlation = corCompSymm(form = ~ Time | id),
		weights = varIdent(form = ~ 1 | Time),
		data=dd1)
summary(fit0)$tTable

fit1 <- gls(Y~-1+time*Trt, 
  correlation = corCompSymm(form = ~ Time | id),
	weights = varIdent(form = ~ 1 | Time),
	data=dd1)

(ctrast <- contrast(fit1, 
  list(time=levels(dd1$time), Trt=levels(dd1$Trt)[2]),
  list(time=levels(dd1$time), Trt=levels(dd1$Trt)[1])))

newdat <- rbind(
  with(contrast(fit1, list(time=levels(dd1$time), Trt=levels(dd1$Trt)[1])),
       data.frame(time = time, Trt = Trt, Estimate = Contrast, SE = SE,
                  Lower = Lower, Upper = Upper, 't-value' = testStat, 'p-value' = Pvalue)
       ),
  with(contrast(fit1, list(time=levels(dd1$time), Trt=levels(dd1$Trt)[2])),
       data.frame(time = time, Trt = Trt, Estimate = Contrast, SE = SE,
                  Lower = Lower, Upper = Upper, 't-value' = testStat, 'p-value' = Pvalue)
       )
  )
newdat$Time <- as.numeric(newdat$time)
ggplot(newdat, aes(y=Estimate, x=Time, color = Trt)) +
    geom_smooth(aes(ymax = Upper, ymin = Lower), stat = 'identity')
```

## Simulate using parallel

```{r, simulations, eval=FALSE}
set.seed(20130318)  
results <- mclapply(1:10000, function(x){
	dd1 <- merge(dd0, data.frame(id = 1:N, drop = sample(1:5, N, replace=TRUE, p=c(0,0.1,0.1,0.1,0.7))))
	dd1$e <- as.numeric(t(rmvnorm(n=N, mean=rep(0,4), sigma=gls_sigma^2*diag(gls_weights)%*%R%*%diag(gls_weights))))
	dd1$Y <- with(dd1, Y0 + e)
	dd1 <- subset(dd1, Time < drop)
	try(summary(update(fit0, data = dd1))$tTable['time4:trt', 'p-value'], silent = TRUE)
}, mc.cores = 10)
```

```{r, load_results, echo=FALSE}
load('2014-04-15-gls-mmrm-power-calcs.Rdata')
```

## Simulated power results

```{r, sim_results}
cat(sum(results < 0.05)/length(results)*100)
```