---
title: "Constrained vs unconstrained longitudinal data analysis"
date: "2016-10-21"
categories: [mixed-effects, lme, nlme, R, constrained longitudinal data analysis, power, simulation, pre-post]
---

# Constrained Longitudinal Data Analysis

[Liang and Zeger (2000)](http://www.jstor.org/stable/25053123) describe a constrained longitudinal data analysis for comparing the time course of two randomized groups in which the model constains the two groups to share a common mean at baseline. [Lu (2010)](https://www.ncbi.nlm.nih.gov/pubmed/19764951) compared the power of this approach to a longitudinal analysis of covariance approach. We conduct a few simple simulation studies to interrogate the effect of the constraint on power to detect group differences. We do not simulate any missing data here.

## Packages used

```{r library, results='hide', message=FALSE}
library(longpower)
library(nlme)
library(mvtnorm)
library(parallel)
library(ggplot2)
library(contrast)
library(tidyr)
library(dplyr)
```

```{r, results='hide', echo=FALSE}
knitr::opts_chunk$set(fig.width=1.7*7, fig.height=7, crop=TRUE)

theme_set(theme_bw(base_size=12))
# The palette with black:
cbbPalette <- c("#0072B2", "#a21933", "#CC79A7", "#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#D55E00")
dxpal <- c("#2b8cbe", "#fc8d59","#e34a33", "#fdcc8a", "#74a9cf")
Palette0 <- Palette <- dxpal
scale_colour_discrete <- function(...) scale_colour_manual(..., values=cbbPalette)
scale_fill_discrete <- function(...) scale_fill_manual(..., values=cbbPalette)
```

# Pre-post design with two repeated measures simulated via random intercept model

## Simulation settings

```{r sim_settings}
sigma_residual <- 1
sigma_random_intercept <- 1
sigma_random_slope <- 0
Time <- 0:1
Beta <- c('(Intercept)'=1, time=0.5, 'trt:time'=0.4)
Beta.null <- c('(Intercept)'=1, time=0.5, 'trt:time'=0)
n <- 200 # per group
N <- 2*n
```

## Analytic calculations

```{r calcs}
cov.t <- function(t1, t2, sig2.i, sig2.s, cov.s.i){
        sig2.i + t1*t2*sig2.s + (t1+t2)*cov.s.i
}
m = length(Time)
R = outer(Time, Time, function(x,y){cov.t(x,y, sigma_random_intercept^2, sigma_random_slope^2, 0)})
R = R + diag(sigma_residual^2, m, m)

power.t.test(n=n, delta=Beta['trt:time'], sd=sqrt(2*sigma_random_intercept))
lmmpower(n=n, delta=Beta['trt:time'], t=Time, 
  sig2.e=sigma_residual^2, sig2.s=sigma_random_slope^2, 
  sig2.i=sigma_random_intercept, cov.s.i=0)
diggle.linear.power(n=n, delta=Beta['trt:time'], t=Time, R=R)
```

## Set up data for simulation

```{r, sim_data_setup}
dd0 <- expand.grid(time = Time, id = 1:N)
dd0$trt <- ifelse(dd0$id <= N/2, 0, 1)
dd0 <- as.data.frame(model.matrix(~ id+trt*time, data = dd0))
dd0$Trt <- as.factor(dd0$trt)
dd0$Y0 <- as.matrix(dd0[, names(Beta)]) %*% Beta
dd0$Y0.null <- as.matrix(dd0[, names(Beta.null)]) %*% Beta.null
ggplot(dd0, aes(x=time, y=Y0, color=Trt)) + geom_line()

set.seed(20130318)
dd1 <- merge(dd0, data.frame(
  id = 1:N, 
  ran.intercept = rnorm(N, sd=sigma_random_intercept),
  ran.slope     = rnorm(N, sd=sigma_random_slope)
))
dd1$e <- rnorm(nrow(dd1), sd=sigma_residual)

dd1$Y <- with(dd1, Y0 + ran.intercept + ran.slope*time + e)
ggplot(dd1, aes(x=time, y=Y, color=Trt, group = id)) + geom_line()

# unconstrained:
fit.LDA <- lme(Y~trt*time, random=~1|id, data=dd1)
printCoefmat(summary(fit.LDA)$tTable)

# constrained:
fit.cLDA <- lme(Y~time+time:trt, random=~1|id, data=dd1)
printCoefmat(summary(fit.cLDA)$tTable)

newdat0 <- expand.grid(trt=0:1, time=Time)

mods <- list(LDA = fit.LDA, cLDA=fit.cLDA)

pd <- do.call(rbind, lapply(names(mods), function(name){
  cbind(newdat0, 
    Y=predict(mods[[name]], newdat0, level=0),
    model = name
  ) 
}))
pd$Trt <- as.factor(pd$trt)
ggplot(pd, aes(y=Y, x=time, color = Trt)) + geom_line() + facet_grid(.~model)
```

## Simulate

```{r, simulations, eval=FALSE}
set.seed(20130318)
results <- do.call(rbind, mclapply(1:10000, function(x){
  dd1 <- merge(dd0, data.frame(
    id = 1:N,
    ran.intercept = rnorm(N, sd=sigma_random_intercept),
    ran.slope     = rnorm(N, sd=sigma_random_slope)
  ))
  dd1$e <- rnorm(nrow(dd1), sd=sigma_residual)
  dd1$Y <- with(dd1, Y0 + ran.intercept + ran.slope*time + e)
  dd1$Y.null <- with(dd1, Y0.null + ran.intercept + ran.slope*time + e)
  ddw <- merge(
    subset(dd1, time==0, c('id', 'trt', 'Y', 'Y.null')),
    subset(dd1, time==1, c('id', 'trt', 'Y', 'Y.null')),
    by=c('id', 'trt'), suffixes=c('.0','.1')
  )
  ddw$d <- with(ddw, Y.1-Y.0)
  ddw$d.null <- with(ddw, Y.null.1-Y.null.0)

  prepost <- list(
    ttest.power   = t.test(         d~trt, data=ddw)$p.value,
    ttest.alpha   = t.test(    d.null~trt, data=ddw)$p.value,
    ancova1.power = summary(lm(     d~Y.0 + trt, data=ddw))$coef['trt', 'Pr(>|t|)'],
    ancova1.alpha = summary(lm(d.null~Y.null.0 + trt, data=ddw))$coef['trt', 'Pr(>|t|)'],
    ancova2.power = summary(lm(     d~scale(Y.0)*trt, data=ddw))$coef['trt', 'Pr(>|t|)'],
    ancova2.alpha = summary(lm(d.null~scale(Y.null.0)*trt, data=ddw))$coef['trt', 'Pr(>|t|)']
  )

  fits <- list(
    LDA.power  = lme(     Y~time*trt, random=~1|id, data=dd1),
    LDA.alpha  = lme(Y.null~time*trt, random=~1|id, data=dd1),
    cLDA.power = lme(     Y~time+time:trt, random=~1|id, data=dd1),
    cLDA.alpha = lme(Y.null~time+time:trt, random=~1|id, data=dd1)
  )

  c(unlist(prepost),
    slopes = unlist(lapply(fits, function(fit){
      summary(fit)$tTable['time:trt', 'p-value']
    })),
    contrasts = unlist(lapply(fits[1:2], function(fit){
      pv <- try(as.numeric(contrast(fit,
        a=list(time=Time[length(Time)], trt=0),
        b=list(time=Time[length(Time)], trt=1))$Pvalue), silent = TRUE)
      if(class(pv) == 'try-error'){
        fit <- update(fit, random=~1|id)
        pv <- try(as.numeric(contrast(fit,
          a=list(time=Time[length(Time)], trt=0),
          b=list(time=Time[length(Time)], trt=1))$Pvalue), silent = TRUE)
      }
      pv
    }))
  )
}, mc.cores = 10))
```

```{r, load_results, echo=FALSE}
load('2016-10-21-constrained-longitudinal-data-analysis-a.Rdata')
```

## Simulated power results (%)

```{r, sim_results}
knitr::kable(cbind('Pr(p<0.05)'=apply(results, 2, function(x){
  sum(x < 0.05)/length(x)*100
})))
```

# Four repeated measures simulated via random slope model
## Simulation settings
```{r}
sigma_residual <- 1
sigma_random_intercept <- 1
sigma_random_slope <- 1
Time <- 0:3
Beta <- c('(Intercept)'=1, time=0.5, 'trt:time'=0.434)
Beta.null <- c('(Intercept)'=1, time=0.5, 'trt:time'=0)
n <- 100 # per group
N <- 2*n
```

## Analytic calculations

```{r}
m = length(Time)
R = outer(Time, Time, function(x,y){cov.t(x,y, sigma_random_intercept^2, sigma_random_slope^2, 0)})
R = R + diag(sigma_residual^2, m, m)

lmmpower(n=n, delta=Beta['trt:time'], t=Time, 
  sig2.e=sigma_residual^2, sig2.s=sigma_random_slope^2, 
  sig2.i=sigma_random_intercept^2, cov.s.i=0)
```

```{r}
diggle.linear.power(n=n, delta=Beta['trt:time'], t=Time, R=R)
```

## Set up data for simulation

```{r}
dd0 <- expand.grid(time = Time, id = 1:N)
dd0$trt <- ifelse(dd0$id <= N/2, 0, 1)
dd0$t <- as.factor(dd0$time)
dd0 <- as.data.frame(model.matrix(~ id+trt*time+t, data = dd0))
dd0$t <- as.factor(dd0$time)
dd0$Trt <- as.factor(dd0$trt)
dd0$Y0 <- as.matrix(dd0[, names(Beta)]) %*% Beta
dd0$Y0.null <- as.matrix(dd0[, names(Beta.null)]) %*% Beta.null

ggplot(dd0, aes(x=time, y=Y0, color=Trt)) + geom_line()
```

```{r}
set.seed(20130318)
dd1 <- merge(dd0, data.frame(
  id = 1:N, 
  ran.intercept = rnorm(N, sd=sigma_random_intercept),
  ran.slope     = rnorm(N, sd=sigma_random_slope)
))
dd1$e <- rnorm(nrow(dd1), sd=sigma_residual)

dd1$Y <- with(dd1, Y0 + ran.intercept + ran.slope*time + e)
ggplot(dd1, aes(x=time, y=Y, color=Trt, group = id)) + geom_line()
```

```{r}
# unconstrained:
fit.LDA <- lme(Y~trt*time, random=~time|id, data=dd1)
printCoefmat(summary(fit.LDA)$tTable)

# constrained:
fit.cLDA <- lme(Y~time+time:trt, random=~time|id, data=dd1)
printCoefmat(summary(fit.cLDA)$tTable)
```

```{r}
newdat0 <- expand.grid(trt=0:1, time=Time)

mods <- list(LDA = fit.LDA, cLDA=fit.cLDA)

pd <- do.call(rbind, lapply(names(mods), function(name){
  cbind(newdat0, 
    Y=predict(mods[[name]], newdat0, level=0),
    model = name
  ) 
}))
pd$Trt <- as.factor(pd$trt)
ggplot(pd, aes(y=Y, x=time, color = Trt)) + geom_line() + facet_grid(.~model)
```

## Simulate

```{r, eval=FALSE}
set.seed(20130318)
results <- do.call(rbind, mclapply(1:10000, function(x){
  dd1 <- merge(dd0, data.frame(
    id = 1:N,
    ran.intercept = rnorm(N, sd=sigma_random_intercept),
    ran.slope     = rnorm(N, sd=sigma_random_slope)
  ))
  dd1$e <- rnorm(nrow(dd1), sd=sigma_residual)
  dd1$Y <- with(dd1, Y0 + ran.intercept + ran.slope*time + e)
  dd1$Y.null <- with(dd1, Y0.null + ran.intercept + ran.slope*time + e)

  fits.cts <- list(
    LDA.power  = lme(     Y~time*trt, random=~time|id, data=dd1, control=lmeControl(returnObject=TRUE)),
    LDA.alpha  = lme(Y.null~time*trt, random=~time|id, data=dd1, control=lmeControl(returnObject=TRUE)),
    cLDA.power = lme(     Y~time+time:trt, random=~time|id, data=dd1, control=lmeControl(returnObject=TRUE)),
    cLDA.alpha = lme(Y.null~time+time:trt, random=~time|id, data=dd1, control=lmeControl(returnObject=TRUE))
  )

  fits.cat <- list(
    LDA.power  = gls(     Y~(t1+t2+t3)*trt, data=dd1,
      correlation = corCompSymm(form = ~ time | id), weights = varIdent(form = ~ 1 | time)),
    LDA.alpha  = gls(Y.null~(t1+t2+t3)*trt, data=dd1,
      correlation = corCompSymm(form = ~ time | id), weights = varIdent(form = ~ 1 | time)),
    cLDA.power = gls(     Y~t1+t2+t3+(t1+t2+t3):trt, data=dd1,
      correlation = corCompSymm(form = ~ time | id), weights = varIdent(form = ~ 1 | time)),
    cLDA.alpha = gls(Y.null~t1+t2+t3+(t1+t2+t3):trt, data=dd1,
      correlation = corCompSymm(form = ~ time | id), weights = varIdent(form = ~ 1 | time))
  )

  c(slopes = unlist(lapply(fits.cts, function(fit){
      summary(fit)$tTable['time:trt', 'p-value']
    })),
    contrasts.cts = unlist(lapply(fits.cts[1:2], function(fit){
      pv <- try(as.numeric(contrast(fit,
        a=list(time=Time[length(Time)], trt=0),
        b=list(time=Time[length(Time)], trt=1))$Pvalue), silent = TRUE)
      if(class(pv) == 'try-error'){
        fit <- update(fit, random=~1|id)
        pv <- try(as.numeric(contrast(fit,
          a=list(time=Time[length(Time)], trt=0),
          b=list(time=Time[length(Time)], trt=1))$Pvalue), silent = TRUE)
      }
      pv
    })),
    contrasts.cat = c(
      unlist(lapply(fits.cat[1:2], function(fit){
        pv <- try(as.numeric(contrast(fit,
          a=list(t1=0, t2=0, t3=1, trt=0),
          b=list(t1=0, t2=0, t3=1, trt=1))$Pvalue), silent = TRUE)
        pv
      })),
      unlist(lapply(fits.cat[3:4], function(fit){
        summary(fit)$tTable['t3:trt', 'p-value']
      }))
    )
  )
}, mc.cores = 10))
```

```{r, load_results_b, echo=FALSE}
load('2016-10-21-constrained-longitudinal-data-analysis-b.Rdata')
```

## Simulated power results (%)
```{r}
knitr::kable(cbind('Pr(p<0.05)'=apply(results, 2, function(x){
  sum(x < 0.05)/length(x)*100
})))
```

# Pre-post design simulated with fixed variance of pre-post difference and various pre-post correlations

## Simulation settings

```{r}
Time <- 0:1
Beta <- c('(Intercept)'=1, time=0.0514, 'trt:time'=0.0514*0.30)
Beta.null <- c('(Intercept)'=1, time=0.0514, 'trt:time'=0)
corr <- c(0.2, 0.5, 0.7)
n <- 64 # per group
N <- 2*n
sd_diff <- 0.0309
```

## Analytic calculations

```{r}
power.t.test(n=n, delta=Beta['trt:time'], sd=sd_diff)
```

## Set up data for simulation

```{r}
dd0 <- expand.grid(time = Time, id = 1:N)
dd0$trt <- ifelse(dd0$id <= N/2, 0, 1)
dd0 <- as.data.frame(model.matrix(~ id+trt*time, data = dd0))
dd0$Trt <- as.factor(dd0$trt)
dd0$Y0 <- as.matrix(dd0[, names(Beta)]) %*% Beta
dd0$Y0.null <- as.matrix(dd0[, names(Beta.null)]) %*% Beta.null
```

## Simulate

```{r, eval=FALSE}
set.seed(20130318)
results <- do.call(rbind, mclapply(1:10000, function(x){
  ll <- lapply(corr, function(rho){
    sd_resid <- sqrt(sd_diff^2/(2*(1-rho)))
    cov_resid <- rho * sd_resid^2
    resids <- gather(
      data.frame(
        id = 1:N,
        rmvnorm(n=N, mean=c(0,0), sigma=matrix(c(sd_resid^2,cov_resid,cov_resid,sd_resid^2), nrow=2))), 
      time, e, 2:3) %>% 
      mutate(time = as.numeric(gsub('X', '', time))-1)
    dd1 <- merge(dd0, resids, by=c('id', 'time'))
    dd1$Y <- with(dd1, Y0 + e)
    dd1$Y.null <- with(dd1, Y0.null + e)
    ddw <- merge(
      subset(dd1, time==0, c('id', 'trt', 'Y', 'Y.null')),
      subset(dd1, time==1, c('id', 'trt', 'Y', 'Y.null')),
      by=c('id', 'trt'), suffixes=c('.0','.1')
    )
    ddw$d <- with(ddw, Y.1-Y.0)
    ddw$d.null <- with(ddw, Y.null.1-Y.null.0)

    prepost <- list(
      ttest.power   = t.test(         d~trt, data=ddw)$p.value,
      ttest.alpha   = t.test(    d.null~trt, data=ddw)$p.value,
      ancova1.power = summary(lm(     d~Y.0 + trt, data=ddw))$coef['trt', 'Pr(>|t|)'],
      ancova1.alpha = summary(lm(d.null~Y.null.0 + trt, data=ddw))$coef['trt', 'Pr(>|t|)'],
      ancova2.power = summary(lm(     d~scale(Y.0)*trt, data=ddw))$coef['trt', 'Pr(>|t|)'],
      ancova2.alpha = summary(lm(d.null~scale(Y.null.0)*trt, data=ddw))$coef['trt', 'Pr(>|t|)']
    )

    fits <- list(
      LDA.power  = lme(     Y~time*trt, random=~1|id, data=dd1, control=lmeControl(returnObject=TRUE)),
      LDA.alpha  = lme(Y.null~time*trt, random=~1|id, data=dd1, control=lmeControl(returnObject=TRUE)),
      cLDA.power = lme(     Y~time+time:trt, random=~1|id, data=dd1, control=lmeControl(returnObject=TRUE)),
      cLDA.alpha = lme(Y.null~time+time:trt, random=~1|id, data=dd1, control=lmeControl(returnObject=TRUE))
    )

    c(unlist(prepost),
      slopes = unlist(lapply(fits, function(fit){
        summary(fit)$tTable['time:trt', 'p-value']
      })),
      contrasts = unlist(lapply(fits[1:2], function(fit){
        pv <- try(as.numeric(contrast(fit,
          a=list(time=Time[length(Time)], trt=0),
          b=list(time=Time[length(Time)], trt=1))$Pvalue), silent = TRUE)
        if(class(pv) == 'try-error'){
          fit <- update(fit, random=~1|id)
          pv <- try(as.numeric(contrast(fit,
            a=list(time=Time[length(Time)], trt=0),
            b=list(time=Time[length(Time)], trt=1))$Pvalue), silent = TRUE)
        }
        pv
      }))
    )
  })
  names(ll) <- corr
  unlist(ll)
}, mc.cores = 10))
```

```{r, load_results_c, echo=FALSE}
load('2016-10-21-constrained-longitudinal-data-analysis-c.Rdata')
```

## Simulated power results (%)

```{r}
knitr::kable(cbind('Pr(p<0.05)'=apply(results, 2, function(x){
  sum(x < 0.05)/length(x)*100
})))
```


So the t-test power is fixed at 80% (by design), the unconstrained LDA has nearly identical power to t-test; while ANCOVA and constrained LDA have similar power (always better power than t-test, but power decreases with increasing pre-post correlation). ANCOVA2 has similar power as ANCOVA1 (not surprising since no interaction between baseline outcome and group was simulated).

This confirms that the cLDA is preferred whenever we have two timepoints in a randomized study; just as ANCOVA is always preferred to t-test. Of course, cLDA has the added advantage of accommodating missing data.
