---
title: "Estimating symptom onset from a single blood test: design artifacts, not biomarker signal"

prefer-html: true
always_allow_html: true
bibliography: references.bib
csl: american-medical-association.csl
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Supplementary Figure S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Supplementary Table S
      space-before-numbering: false
      latex-list-of-description: Supplementary Table
format:
  html:
    code-fold: true
    toc: true
  pdf:
    documentclass: article
    geometry:
      - margin=1in
  latex:
    documentclass: article
    geometry:
      - margin=1in
  docx: default
---

**Philip S. Insel**^1^\* and **Michael C. Donohue**^2^

^1^Department of Psychiatry and Behavioral Sciences, University of California, San Francisco, San Francisco, CA, USA

^2^USC Epstein Family Alzheimer's Therapeutic Research Institute, University of Southern California, San Diego, CA, USA

\*Corresponding author: philip.insel@ucsf.edu

\newpage

Petersen et al.[@petersen2026] report that disease-clock models applied to plasma %p-tau217 can estimate the age at symptom onset of Alzheimer's disease, concluding that "the time until onset of AD symptoms can be estimated using a single blood test within a margin of error that is acceptable for use in clinical trials," on the basis of adjusted R² values of 0.337–0.612. However, their constructed predictor, the clock-derived estimate of the age at plasma %p-tau217 positivity, combines chronological age with information from a blood test, and the reported performance is misattributed to the blood test. Using their data, we show that this performance is almost entirely driven by study design rather than biomarker signal: the blood-test component alone explains essentially none of the variance in age at symptom onset (R² = 0.01–0.02). Indeed, high R² values attributable to chronological age alone are guaranteed by the design. They are obtained even when the blood-test information is replaced with random numbers, demonstrating that the apparent performance requires no biomarker signal at all. Moreover, the clock generally performs worse than a simple model including chronological age and raw %p-tau217 directly. The conclusion that the time until AD symptoms can be estimated from a single blood test, within a margin acceptable for clinical trials, is unfounded.

Plasma markers of phosphorylated tau-217 (such as %p-tau217) have rapidly emerged as robust markers of underlying Alzheimer's disease (AD) pathology.[@warmenhoven2025comprehensive] Recent work has extended this to longitudinal measurements, which show p-tau217 changes over time in at-risk groups,[@mattsson2020longitudinal] and p-tau217 associations with future cognitive decline.[@insel2025concurrent] Petersen et al. build on this by applying disease-clock models (SILA[@betthauser2022multi] and TIRA[@schindler2021predicting]) to estimate each individual's age at plasma %p-tau217 positivity, defined as baseline age minus the estimated time from positivity, and then regress observed age at symptom onset on this estimate in individuals who progressed to symptomatic AD.

This design produces an artifact for two reasons. First, restricting the analysis to progressors within a limited follow-up window bounds the interval between baseline age and symptom onset, so baseline age alone strongly predicts onset age even with no biomarker information. Second, the clock predictor is itself constructed from baseline age, so it shares that same age component with the outcome, making the resulting association partly self-referential. More generally, this artifact arises whenever the outcome is constrained by an earlier time variable and the predictor contains that same or a related time component, inducing associations through shared structure rather than independent signal. To separate the contribution of chronological age, which is confounded with the design artifact, from real biomarker signal, we used the data from Petersen et al. to decompose the clock into baseline age and estimated time from %p-tau217 positivity. Symptom onset was then modeled as a function of both terms. As a null comparison, we replaced the biomarker timing component with a random value over the same range. We further compared the clock against a model using baseline age and raw %p-tau217 level directly.

When the clock is decomposed into baseline age and estimated time from %p-tau217 positivity, the biomarker timing term alone explained negligible variance in age at symptom onset in either cohort: adjusted R² = 0.02 (TIRA) and 0.01 (SILA) in the primary Knight ADRC cohort (@fig-tira-sila-kadrc), and essentially zero in ADNI (@suppfig-tira-sila-adni; in fact slightly negative for both methods). Baseline age alone explained far more: R² = 0.54 for both method samples in Knight ADRC, and R² = 0.71 (TIRA sample) and 0.61 (SILA sample) in ADNI. This association is imposed entirely by the study design rather than reflecting biomarker signal, as shown by the pink regions in @fig-tira-sila-kadrc: they define the region the data must occupy, given that symptom onset occurs between 5.7 years before and 16.4 years after baseline. Because onset must fall within this bounded window, baseline age alone determines age at symptom onset up to a fixed interval, forcing the strong positive association seen in the left column regardless of any biomarker information.

Incorporating the timing term through the clock produced a small improvement over age alone in Knight ADRC (R² = 0.60–0.61, versus 0.54 for age alone), but this improvement was small (ΔR² = 0.07) and did not replicate in the validation cohort, indicating instability rather than robust incremental value. In ADNI, incorporating the biomarker through the clock did not add information but instead injected noise, lowering performance below age alone and below a random comparator: for TIRA, the clock explained R² = 0.34 versus 0.71 for age alone, and for SILA, R² = 0.47 versus 0.61, compared with R² = 0.56 for the random comparator.


```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(yhat)
library(flextable)
library(knitr)
library(ggtext)
library(officer)

source('tira-sila-helper-functions.R')

# library(tidyverse)
# library(fuzzyjoin)
# library(gtsummary)
# library(yhat)
# library(magrittr)

options(htmltools.dir.version = FALSE, knitr.kable.NA = '')

opts_chunk$set(
  echo = FALSE,
  tidy=FALSE,
  collapse = TRUE,
  fig.path = '',
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  message = FALSE,
  warning = FALSE, 
  out.extra = '',
  out.width='\\linewidth',
  fig.align = 'center', 
  crop = TRUE, 
  fig.pos = '!h',
  cache=TRUE,
  comment = '',
  fig.dpi = 300)
```

```{r ggplot-setup}
theme_jama <- function(base_size = 10) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # Overall plot settings
      plot.title = element_text(size = base_size + 1, face = "bold", 
        hjust = 0, margin = margin(b = 10)),
      plot.subtitle = element_text(size = base_size, hjust = 0, 
        margin = margin(b = 10)),
      plot.caption = element_text(size = base_size - 1, hjust = 1, 
        margin = margin(t = 10)),
      
      # Axis settings
      axis.title = element_text(size = base_size, face = "bold"),
      axis.text = element_text(size = base_size - 1, color = "black"),
      axis.line = element_line(color = "black", linewidth = 0.5),
      axis.ticks = element_line(color = "black", linewidth = 0.5),
      
      # Panel settings
      panel.background = element_rect(fill = "white", color = NA),
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      
      # Legend settings
      legend.title = element_text(size = base_size, face = "bold"),
      legend.text = element_text(size = base_size - 1),
      legend.key = element_rect(fill = "white", color = NA),
      legend.background = element_rect(fill = "transparent", color = NA),
      legend.position = "bottom",
      
      # Strip settings (for facets)
      strip.background = element_blank(),
      strip.text = element_text(size = base_size - 2, face = "bold",
        vjust = 1, color = "black"),
      
      # Remove plot background
      plot.background = element_rect(fill = "white", color = NA)
    )
}

# JAMA-appropriate color palette (colorblind-friendly)
jama_colors <- c("#374E55", "#DF8F44", "#00A1D5", "#B24745", 
  "#79AF97", "#6A6599", "#80796B")

# Scale functions for easy use
scale_color_jama <- function(...) {
  scale_color_manual(values = jama_colors, ...)
}

scale_fill_jama <- function(...) {
  scale_fill_manual(values = jama_colors, ...)
}

# Example usage:
# ggplot(data, aes(x = x, y = y, color = group)) +
#   geom_point() +
#   theme_jama() +
#   scale_color_jama()
```

```{r merge-provided-data, eval = FALSE}
adni_tira <- read.csv("Petersen progressor subsets/progressors_ADNI_TIRA_anon.csv")
adni_sila <- read.csv("Petersen progressor subsets/progressors_ADNI_SILA_anon.csv")
kadrc_tira <- read.csv("Petersen progressor subsets/progressors_KADRC_TIRA_anon.csv")
kadrc_sila <- read.csv("Petersen progressor subsets/progressors_KADRC_SILA_anon.csv")

dat <- adni_tira %>% mutate(Source = 'ADNI', Method = 'TIRA') %>%
  bind_rows(adni_sila %>% mutate(Source = 'ADNI', Method = 'SILA')) %>%
  bind_rows(kadrc_tira %>% mutate(Source = 'Knight ADRC', Method = 'TIRA')) %>%
  bind_rows(kadrc_sila %>% mutate(Source = 'Knight ADRC', Method = 'SILA')) %>%
  # ptau217_positivity_age = baseline_plasma_age - time_from_positivity
  mutate(
    time_from_positivity = baseline_plasma_age - ptau217_positivity_age,
    time_from_onset = baseline_plasma_age - age_of_symptom_onset,
    id = paste(Source, baseline_plasma_age, baseline_plasma) %>%
      as.factor() %>% as.numeric())

if(with(dat, any(duplicated(paste(Method, id))))){
  stop("IDs are not unique.")
}

# add random ----
max_pos <- random_time_from_positivity <- dat %>%
  filter(Method == "SILA") %>%
  pull(time_from_positivity) %>%
  max()

set.seed(20260317)
random_time_from_positivity <- dat %>%
  filter(Method == "SILA" | id == 53) %>%
  arrange(Source, Method) %>%
  mutate(
    Method = "Random",
    time_from_positivity = runif(n(), min = time_from_onset, max = max_pos),
    ptau217_positivity_age = baseline_plasma_age - time_from_positivity)

# ggplot(random_time_from_positivity,
#   aes(x=baseline_plasma_age, y=ptau217_positivity_age)) +
#   geom_point() +
#   geom_abline(a=0, b=1, linetype='dashed')
# 
# ggplot(random_time_from_positivity,
#   aes(x=ptau217_positivity_age, y=age_of_symptom_onset)) +
#   geom_point() +
#   geom_abline(a=0, b=1, linetype='dashed')

dat <- dat %>%
  bind_rows(random_time_from_positivity) %>%
  mutate(
    Source = factor(Source, c('ADNI', 'Knight ADRC')),
    Method = factor(Method, c('TIRA', 'SILA', 'Random')),
    time_from_positivity_to_onset = age_of_symptom_onset - ptau217_positivity_age)

# ggplot(dat, aes(x = baseline_plasma_age, y = ptau217_positivity_age)) +
#   geom_point() +
#   facet_grid(Source ~ Method) +
#   geom_abline(a = 0, b = 1, linetype = 'dashed')
# 
# ggplot(dat, aes(x = ptau217_positivity_age, y = age_of_symptom_onset)) +
#   geom_point() +
#   facet_grid(Source ~ Method) +
#   geom_abline(a = 0, b = 1, linetype = 'dashed')

save(dat, file = 'petersen-tira-sila.rdata')
```

```{r load-scraped-data}
load('petersen-tira-sila.rdata')

max_pos <- random_time_from_positivity <- dat %>%
  filter(Method == "SILA") %>%
  pull(time_from_positivity) %>%
  max()

dat_long <- dat %>%
  pivot_longer(
    cols = c(baseline_plasma_age, time_from_positivity, 
      ptau217_positivity_age, baseline_plasma),
    names_to = "panel",
    values_to = "x") %>%
  mutate(
    y = age_of_symptom_onset,
    panel = factor(
      panel,
      levels = c("baseline_plasma_age", "time_from_positivity",
        "ptau217_positivity_age"),
      labels = c("Baseline age (yrs)", 
        "Estimated time from %p-tau217 positivity (yrs)", 
        "Estimated age at %p-tau217 positivity (yrs)")))
```

```{r panel-stats}
panel_stats_adni  <- compute_panel_stats(dat_long, "ADNI")
panel_stats_kadrc <- compute_panel_stats(dat_long, "Knight ADRC")
seg_data_adni  <- make_seg_data(dat, "ADNI", levels(dat_long$panel))
seg_data_kadrc <- make_seg_data(dat, "Knight ADRC", levels(dat_long$panel))
constraint_adni  <- make_constraint(dat_long, "ADNI")
constraint_kadrc <- make_constraint(dat_long, "Knight ADRC")
```

```{r fig-tira-sila-kadrc, fig.width=8.5, fig.height=8.5}
#| fig-cap: "Relationships in Knight ADRC between the predicted outcome (age of symptom onset) and three predictors: age at baseline (first plasma visit), estimated time from %p-tau217 positivity, and their difference: baseline age minus estimated time from %p-tau217 positivity (i.e., age of %p-tau217 positivity). Time from %p-tau217 positivity is estimated by the TIRA method (top row), the SILA method (middle row), or by randomly sampling from a uniform distribution over the observed range of estimates (bottom row). Panels include fitted linear regression estimates, summary statistics for each predictor, and a dashed identity line. To illustrate the contribution of time from %p-tau217 positivity (middle column), the rightmost panels show faint gray lines connecting each person's age at baseline to their estimated age of %p-tau217 positivity. The pink regions span the observed window of onset: between 5.7 years before and 16.4 years after baseline."

plot_panels(dat_long, "Knight ADRC", panel_stats_kadrc, seg_data_kadrc, constraint_kadrc, y_pos_override = 120)
```

These data do not suggest that plasma %p-tau217 lacks predictive value; a substantial literature establishes that it is strongly associated with cognitive decline. [@Sperling2024; @insel2025concurrent] What these data show is that the disease clock does not extract additional prognostic information from the biomarker. It degrades information that is already available in the raw measurement. When %p-tau217 was entered into the model directly, as a raw level alongside baseline age, it outperformed the clock-derived predictor in every cohort and clock method (@fig-model-R2). The clock fared worst in the validation cohort, where it fell well below direct inclusion of the raw measurement and, for both methods, was considerably worse than a predictor in which the timing term was replaced with a random number. The same biomarker information that improves prediction when used directly is distorted by the clock measure.

```{r fig-model-R2, fig.width=8.5, fig.height=8.5}
#| fig-cap: "**Model comparison across cohorts and clock methods.** Adjusted R² for four models of age at symptom onset in ADNI (top row) and Knight ADRC (bottom row): baseline age alone (dark), the clock-derived age at %p-tau217 positivity (orange in the SILA column, blue in the TIRA column), a random comparator for age at %p-tau217 positivity drawn uniformly over the observed range (grey), and baseline age together with raw %p-tau217 level (red). Columns correspond to the samples fit with SILA and TIRA. Values above bars show the change in Akaike information criterion (ΔAIC) relative to the baseline-age model in the same panel; negative values indicate improved fit and positive values indicate worse fit. Across cohorts, the clock-derived predictor does not improve on direct inclusion of age and raw %p-tau217."

# Build results table ----
tab <- NULL

for (src in c("ADNI", "Knight ADRC")) {
  for (meth in c("SILA", "TIRA")) {
    
    df_sub <- dat %>% filter(Source == src, Method == meth)
    df_sub_rand <- dat %>% filter(Source == src, Method == "Random") %>%
      filter(id %in% df_sub$id)
    if(nrow(df_sub) != nrow(df_sub_rand)){
      stop("random time data is different size")
    }
    
    bl_plasma_age <- lm(age_of_symptom_onset ~ baseline_plasma_age, 
      data = df_sub)
    ptau217_positivity_age <- lm(age_of_symptom_onset ~ ptau217_positivity_age, 
      data = df_sub)
    rand_ptau217_positivity_age <- lm(age_of_symptom_onset ~ ptau217_positivity_age, 
      data = df_sub_rand)
    bl_plasma_age_ptau217 <- lm(age_of_symptom_onset ~ baseline_plasma_age + 
        baseline_plasma, data = df_sub)
    
    mods <- list(
      "Baseline age" = bl_plasma_age,
      "Age of %p-tau217 positivity" = ptau217_positivity_age,
      "Random age of %p-tau217 positivity" = rand_ptau217_positivity_age,
      "Baseline age + %p-tau217" = bl_plasma_age_ptau217
    )
    
    AIC0 <- AIC(bl_plasma_age)
    
    tab <- tab %>% bind_rows(tibble(
      Source = src,
      Method = meth,
      Model = names(mods),
      R2 = unlist(lapply(mods, function(x) summary(x)$r.squared)),
      Adj_R2 = unlist(lapply(mods, function(x) summary(x)$adj.r.squared)),
      AIC = unlist(lapply(mods, function(x) AIC(x))),
      delta_AIC = unlist(lapply(mods, function(x) AIC(x))) - AIC0
    ))
  }
}

# Set factor order for nicer plotting ----
tab <- tab %>%
  mutate(
    Model.x = factor(Model, levels = c(
      "Baseline age",
      "Age of %p-tau217 positivity",
      "Random age of %p-tau217 positivity",
      "Baseline age + %p-tau217")),
    Sample = factor(Method, 
      levels = c("SILA", "TIRA"), labels = c("SILA Sample", "TIRA Sample")),
    Model = case_when(
      Model.x == "Age of %p-tau217 positivity" ~ as.character(Method),
      TRUE ~ as.character(Model.x)) %>%
      factor(c("Baseline age", "SILA", "TIRA", "Random", "Baseline age + %p-tau217")),
    AIC.label = case_when(
      Model.x == "Baseline age" ~ '',
      TRUE ~ paste0("(", sprintf("%.1f", delta_AIC), ")")))

# Plot
p <- ggplot(tab, aes(x = Model.x, y = Adj_R2, fill = Model)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  # Adj R2 labels (inside bars)
  geom_text(
    aes(label = sprintf("%.2f", Adj_R2)),
    vjust = 1.3,
    color = "white",
    size = 5) +
  # Delta AIC labels (above bars)
  geom_text(
    aes(label = AIC.label),
    vjust = -0.5,
    size = 5) +
  facet_grid(Source ~ Sample) +
  scale_y_continuous(
    limits = c(0, max(tab$Adj_R2, na.rm = TRUE) * 1.15),
    expand = expansion(mult = c(0, 0.05))) +
  labs(
    x = NULL,
    y = "Adjusted R²") +
  theme_jama() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 30, hjust = 1, size = 10),
    axis.title.y = element_text(size = 15),
    strip.text.x = element_text(size = 15),
    strip.text.y = element_text(size = 15),
    strip.placement = "outside",
    strip.clip = "off",
    panel.spacing = unit(2, "lines")) +
  scale_fill_manual(values = jama_colors)

p
```

The same structural mechanism operates in a second analysis presented in the paper, in which time from %p-tau217 positivity to symptom onset is regressed on the estimated age at %p-tau217 positivity (their Supplementary Fig. 6), an analysis in which age does not appear in the outcome. This relocates the structural constraint from baseline age to age at %p-tau217 positivity rather than removing it. Because observed symptom onset in these cohorts falls between roughly 68 and 98 years, the duration is bounded by the age at positivity. An individual estimated to become positive at age 90 can have a duration of at most about 8 years, since onset is essentially never observed beyond 98, whereas an individual estimated to become positive at age 60 must have a duration of at least about 8 years, since onset is essentially never observed before 68. As in the primary analysis, this bound is shown directly by the shaded regions in @fig-supp-fig-6, which mark the range the data are constrained to occupy given the observed span of onset ages. Replacing the clock-derived estimate with baseline age minus a randomly generated time reproduced a negative association of similar magnitude to the SILA and TIRA estimates, confirming that the relationship is driven by the structural dependence of duration on age at positivity rather than by biomarker-derived timing. The authors conclude from this analysis that "the time from %p-tau217 positivity to onset of AD symptoms was markedly shorter in older individuals." This relationship is guaranteed by the bound and predetermined by the design, arising even in the absence of any underlying biological relationship. This second conclusion of the paper is therefore not supported.

```{r fig-supp-fig-6, fig.width=8.5, fig.height=8.5}
#| fig-cap: "Relationships between age of symptom onset minus age of %p-tau217 positivity versus age of positivity (TIRA, SILA, or Random). Panels include fitted linear regression estimates, and summary statistics for each. The pink shaded regions show the range of possible values given the observed ages of symptom onset, which span from 68 to 98 years."

age_pos_range <- filter(dat_long, Source == 'Knight ADRC', 
  panel == 'Estimated age at %p-tau217 positivity (yrs)') %>% 
  pull(x) %>% range() %>% round()

age_ons_range <- filter(dat_long, Source == 'Knight ADRC', 
  panel == 'Estimated age at %p-tau217 positivity (yrs)') %>% 
  pull(y) %>% range() %>% round()

constraint_ribbon_kadrc <- 
  expand_grid(
    x = age_pos_range[1]:age_pos_range[2], 
    Method = unique(dat_long$Method)) %>%
  mutate(
    Source = 'Knight ADRC',
    y_min = age_ons_range[1] - x, 
    y_max = age_ons_range[2] - x, time_from_positivity_to_onset = NA_real_)

age_pos_range <- filter(dat_long, Source == 'ADNI', 
  panel == 'Estimated age at %p-tau217 positivity (yrs)') %>% 
  pull(x) %>% range() %>% round()

constraint_ribbon_adni <- 
  expand_grid(
    x = age_pos_range[1]:age_pos_range[2], 
    Method = unique(dat_long$Method)) %>%
  mutate(
    Source = 'ADNI',
    y_min = age_ons_range[1] - x, 
    y_max = age_ons_range[2] - x, time_from_positivity_to_onset = NA_real_)

constraint_ribbon <- bind_rows(constraint_ribbon_kadrc, constraint_ribbon_adni)

# p1_stats----
p1_stats <- dat %>%
  mutate(x = ptau217_positivity_age, y = time_from_positivity_to_onset) %>%
  group_by(Source, Method) %>%
  summarise(
    N   = sum(complete.cases(x, y)),
    rho = cor(x, y, method = "spearman", use = "complete.obs"),
    fit = list(lm(y ~ x)),
    beta   = sapply(fit, function(m) coef(m)[2]),
    alpha  = sapply(fit, function(m) coef(m)[1]),
    adjR2  = sapply(fit, function(m) summary(m)$adj.r.squared),
    x_min = min(x, na.rm = TRUE),    
    x_max = max(x, na.rm = TRUE),    
    y_min = 0,    
    y_max = 30) %>%
  mutate(row_id = row_number()) %>%
  ungroup() %>%
  mutate(
    label  = sprintf(
      "y = %.2fx + %.1f <br>Adjusted R<sup>2</sup>=%.2f<br>Spearman \u03c1 = %.2f<br>N = %d",
      beta, alpha, adjR2, rho, N),
    # place text in upper left
    x_range = x_max - x_min,
    y_range = y_max - y_min,
    x_pos = x_max - 0.7 * x_range,                           # a bit in from left
    y_pos = y_max - 0.05 * y_range             # spaced downward
  )

p1_stats$y_pos <- 45

ggplot(dat, aes(x = ptau217_positivity_age, y = time_from_positivity_to_onset)) +
  geom_point(size = 2.5, alpha = 0.4) +
  geom_smooth(method = 'lm') +
  facet_grid(cols=vars(Method), rows=vars(Source)) +
  geom_richtext(
    data = p1_stats,
    aes(x = x_pos, y = y_pos, label = label),
    hjust = 0,           # left aligned
    vjust = 1,           # top of text block
    lineheight = 1.1,
    size = 3.5,
    fill = NA, label.color = NA,
    show.legend = FALSE) +  # text does not need legend entry
  geom_ribbon(data = constraint_ribbon, aes(x=x, ymin = y_min, ymax = y_max),
    alpha = 0.3, fill = "pink") +
  ylab("Time from positivity to symptom onset (yrs)") +
  xlab("Estimated age at plasma %p-tau217 positivity (yrs)") +
  theme_jama() +
  theme(
    strip.text = element_text(size = 15))
```

The event-time summaries in Petersen et al. (their Figure 3b), which compare time from %p-tau217 positivity to symptom onset across groups defined by age at positivity, are a grouped form of the same duration outcome and are subject to the same constraint. Although the underlying survival analysis appropriately includes nonprogressors and censoring, individuals who become positive at older ages necessarily have less observable time before onset, so the shorter durations seen at older positivity ages partly reflect this structural bound rather than faster disease progression.

Taken together, these analyses demonstrate that the clock-based predictor inherits structural dependence from shared age components and bounded follow-up, producing apparently strong associations with symptom onset even in the absence of biomarker timing information. Because the outcome is constrained by age and the predictor is built from age, the reported R² is largely a product of the design rather than of biomarker signal: it is obtained even when the biomarker component is replaced with a random value, as the random comparator shows throughout. The reported associations can therefore appear strong and biologically meaningful while reflecting little independent biomarker information. Separately, even where the biomarker does carry signal, there is a loss of information under the clock formulation, demonstrated by its poorer predictive performance relative to the raw %p-tau217 level. This loss follows from the clock's central assumption that all individuals follow a common accumulation trajectory, so that time from positivity can be read from a group-level mean given a single measurement — yet individuals do not progress along a shared trajectory, and the subject-level heterogeneity this assumption discards is precisely the information needed to estimate individual timing. These issues are not specific to plasma biomarkers: the same structural features and modeling assumptions arise in amyloid PET clock models[@schindler2021predicting] that relate estimated age at positivity to symptom onset, and such disease-time models are increasingly common, underscoring the need to scrutinize their statistical properties and performance relative to more conventional approaches.

Decomposed, the central claim — that the information in a single blood test can predict the age of symptom onset with clinical trial-level accuracy — collapses. The R² attributed to the blood test was 0.337 to 0.612, but the R² for the blood test alone is 0.01 to 0.02. Establishing genuine predictive value requires performance that exceeds what chronological age and the structure of the analysis already guarantee. Such claims are further inflated by evaluating only the minority who progress while ignoring the stable majority, an approach that does not reflect the clinical reality of population screening. Given the clear clinical potential of plasma measures of AD, and the growing availability of direct-to-consumer p-tau217 testing, it is critical that research evaluating these tools accurately represents what they can and cannot provide. Rather than assuming that all individuals follow a common declining trajectory, approaches that explicitly estimate the probability that an individual will progress convey the uncertainty in that individual's course — which is clinically essential before counseling anyone on a "clock" premised on inevitable decline.

\clearpage

# Data and code availability

See code at GitHub: <https://github.com/mcdonohue/blog/tree/main/posts/design-artifacts>. Data can be requested directly from Schindler, et al.

\newpage

# References {.unnumbered}

::: {#refs}
:::

\newpage

# Supplementary Information

::: {#suppfig-tira-sila-adni}

```{r suppfig-tira-sila-adni}
#| echo: false
#| fig-width: 8.5
#| fig-height: 8.5

plot_panels(dat_long, "ADNI", panel_stats_adni, seg_data_adni, constraint_adni, y_pos_override = 99)
```

Relationships in ADNI between the predicted outcome (age of symptom onset) and three predictors: age at baseline (first plasma visit), estimated time from %p-tau217 positivity, and their difference: baseline age minus estimated time from %p-tau217 positivity (i.e., age of %p-tau217 positivity). Time from %p-tau217 positivity is estimated by the TIRA method (top row), the SILA method (middle row), or by randomly sampling from a uniform distribution over the observed range of estimates (bottom row). Panels include fitted linear regression estimates, summary statistics for each predictor, and a dashed identity line. To illustrate the contribution of time from %p-tau217 positivity (middle column), the rightmost panels show faint gray lines connecting each person's age at baseline to their estimated age of %p-tau217 positivity. The pink regions span the observed window of onset ages: between 2.1 years before and 10.3 years after baseline.

:::
