---
title: "GLS, MMRM, and Missing Data"
date: "2013-05-06"
categories: [GLS, MMRM, mixed-effects, lme, nlme, R]
summary: Fitting MMRM in R with missing data.
---

# The issue

The `gls` routine in the `nlme` package is sensitive to the data order when using `corCompSymm`, particularly when data is missing. This is alluded to in the `corCompSymm` documentation. Below are some examples. 

## No missingness

When there is no missingness, there is no problem provided the data is sorted by time:

```{r, echo = FALSE, results = 'hide'}
library(ggplot2)

knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width=1.7*7, fig.height=7, crop=TRUE,
  tidy = FALSE, comment = NA, cache = FALSE)

theme_set(theme_bw(base_size=12))
# The palette with black:
cbbPalette <- c("#0072B2", "#a21933", "#CC79A7", "#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#D55E00")
dxpal <- c("#2b8cbe", "#fc8d59","#e34a33", "#fdcc8a", "#74a9cf")
Palette0 <- Palette <- dxpal
scale_colour_discrete <- function(...) scale_colour_manual(..., values=cbbPalette)
scale_fill_discrete <- function(...) scale_fill_manual(..., values=cbbPalette)
```

```{r}
library(nlme)
set.seed(20130506)
fm1Orth.gls <- gls( distance ~ Sex * I(age - 11), 
  Orthodont,
  correlation = corSymm(form = ~ 1 | Subject),
  weights = varIdent(form = ~ 1 | age) )
summary(fm1Orth.gls)$tTable

fm2Orth.gls <- gls( distance ~ Sex * I(age - 11), 
  Orthodont[nrow(Orthodont):1, ], # reverse order
  correlation = corSymm(form = ~ 1 | Subject),
  weights = varIdent(form = ~ 1 | age) )
summary(fm2Orth.gls)$tTable # same as sorted
```

But if the data is not sorted by time, we get inconsistent results. In fact, here the estimation does not converge:

```{r, eval = FALSE}
fm3Orth.gls <- gls( distance ~ Sex * I(age - 11), 
  Orthodont[sample(1:nrow(Orthodont), replace = TRUE), ],
  correlation = corSymm(form = ~ 1 | Subject),
  weights = varIdent(form = ~ 1 | age) )
```

## Missingness
When there is missingness, sorting does not help:
```{r}
# Generate some missingness
Orthodont$missing <- sample(c(TRUE, FALSE), 
  size = nrow(Orthodont), prob = c(0.2, 0.8), 
  replace = TRUE)

fm4Orth.gls <- gls( distance ~ Sex * I(age - 11), 
  Orthodont, subset = !missing,
  correlation = corSymm(form = ~ 1 | Subject),
  weights = varIdent(form = ~ 1 | age) )
summary(fm4Orth.gls)$tTable

fm5Orth.gls <- gls( distance ~ Sex * I(age - 11), 
  Orthodont[nrow(Orthodont):1, ], # reverse order
  subset = !missing,
  correlation = corSymm(form = ~ 1 | Subject),
  weights = varIdent(form = ~ 1 | age) )
summary(fm5Orth.gls)$tTable
```

# The fix
By default, corSymm(form = ~ 1 | Subject) uses the order of observations to estimate the correlation components. In this case, the (i,j) component of the correlation matrix represents the correlation between the ith and jth observation. If instead we want the correlation between the ith and jth time point to be represented, we need to use corSymm(form = ~ t | Subject), where t are the consecutive integers indexing time.

```{r}
Orthodont$t <- as.numeric(factor(Orthodont$age, levels = c(8, 10, 12, 14)))
with(Orthodont, table(t, age))
fm6Orth.gls <- gls( distance ~ Sex * I(age - 11), 
  Orthodont, subset = !missing,
  correlation = corSymm(form = ~ t | Subject),
  weights = varIdent(form = ~ 1 | age) )
summary(fm6Orth.gls)$tTable

fm7Orth.gls <- gls( distance ~ Sex * I(age - 11), 
  Orthodont[nrow(Orthodont):1, ], 
  subset = !missing,
  correlation = corSymm(form = ~ t | Subject),
  weights = varIdent(form = ~ 1 | age) )
summary(fm7Orth.gls)$tTable
```

Note this is not an issue with compound symmetry structure, corCompSymm, because the correlation parameters are exchangeable with respect to time. 
